{DEM}-{SPH} study of particles dispersion in fluid

### 2024-05-31 manuscript/extended_abstract.tex
Add an extended abstract

\documentclass[preprint,12pt]{elsarticle}
% \documentclass[preprint,12pt,draft]{elsarticle}
% \documentclass[draft,12pt]{elsarticle}

\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{multirow}
\usepackage{relsize}
\usepackage[utf8]{inputenc}
\usepackage[capitalise]{cleveref}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\usepackage[section]{placeins}
\usepackage{booktabs}
\usepackage{url}

@@

\usepackage{lineno}
% \linenumbers

\journal{}

@@

\begin{frontmatter}

  \title{{DEM}-{SPH} study of particles dispersion in fluid}
  \author[xxx]{Dinesh Adepu\corref{cor1}}
  \ead{d.dinesh@surrey.ac.uk}
  \author[xxx]{Chuan Yu Wu}
  % \ead{XXX}
\address[xxx]{School of Chemistry and Chemical Engineering, University of Surrey}

\cortext[cor1]{Corresponding author}

\begin{abstract}
  Mixing powdered substances in tanks using a stirrer is a common occurrence
  across various industries. Typically, this phenomenon is studied numerically
  due to the complex nonlinear physics involved. In our current study, we have
  developed a solver combining Smoothed Particle Hydrodynamics (SPH) and
  Discrete Element Method (DEM) to investigate the mixing behavior of rigid
  circular particles within a tank under the influence of a stirrer operating
  at different speeds.  The SPH method governs the fluid phase, while the
  dynamics and interactions of particles are captured by the DEM. We achieve a
  fully resolved coupling between solid particles and fluid particles by
  discretizing solid particles into dummy SPH particles. Our validation
  process includes verifying the fluid solver's accuracy through a Poiseuille
  problem and validating the DEM solver by benchmarking interactions between
  two particles and particle-wall impacts.  The coupled model's is validated
  by simulating single-particle entry in a 2D steady tank and a cube settling
  in a tank, which is a 3D case. Following validation, we proceed with the
  mixing study. We conduct two simulations: one with spherical particles of
  uniform radius and another with particles having two different radius.  Our
  findings reveal that as for lower stirrer speeds, particles tend to
  aggregate initially and remain in the center. However, with further increase
  in stirrer speed, particles tend to accumulate near the tank corners due to
  the circulation induced in the fluid due to the stirrer motion. We use an
  open-source software, PySPH, adapting and integrating the coupled model to
  conduct the current study.
\end{abstract}

\begin{keyword}
%% keywords here, in the form: keyword \sep keyword
{particle dispersion}, {particle mixing}, {SPH-DEM}, {stirrer}

%% MSC codes here, in the form: \MSC code \sep code
%% or \MSC[2008] code \sep code (2000 is the default)

\end{keyword}

\end{frontmatter}

% \linenumbers

\FloatBarrier%
\section{Introduction}

Particle dispersion is a fundamental physical phenomenon encountered in
various industries where powder and fluid interactions are pivotal.  Petroleum
industry \cite{wang2023developments}, the transport of bodies in internal
systems \cite{Dai2021}, debris flow \cite{Qingyun2022}, the food processing
industry \cite{Karunasena2014}, and ice-sea modeling \cite{Mintu2018} are a
few areas to mention.  In many processes the mixing of powder with fluid using
a stirrer are commonplace \cite{li2022study}.  \Cref{fig:gloabl_body_frame_rb}
illustrates a sample particle dispersion problem studied in the current work.
While experimental investigation of such processes is often impractical due to
their highly nonlinear nature, numerical methods offer a viable alternative.
While these systems are studied with numerical methods, they are a part of
two-way coupling models. In numerical method approach, one can employ either a
mesh-based or meshless technique to study the current phenomena.

\begin{figure}[!htpb]
  \centering
  \includegraphics[width=0.4\textwidth]{images/hs_tank_fluid_with_particles}
  \caption{An illustrative figure showcasing the dispersion of particles within a free surface tank.}
  \label{intro:schematic}
\end{figure}

The combination of Discrete Element Method (DEM) for solid particle
interactions and Computational Fluid Dynamics (CFD) for fluid modeling
is a classic approach in handling particulate flow problem. In
mesh-based modeling, lattice Boltzmann method (LBM)
\cite{xiong2014lbm} and finite volume method (FVM) \cite{kloss2012models} are
utilized to handle the fluid dynamics. Similarly, in meshless schemes like
Smoothed Particle Hydrodynamics (SPH) \cite{peng2021fully},
Moving Particle Semi-implicit (MPS), and Particle Finite Element
Method (PFEM) \cite{li2019modeling, franci2020pfem}, handles the fluid
dynamics.

While handling the interaction between the fluid and solid particles, the
interaction can be categorized into a resolved and unresolved coupling.  In
unresolved coupling, simplified models are utilized to calculate the forces
exerted on solid particles within the fluid flow. In contrast, in the fully
resolved coupling approach, the fluid forces are directly computed on the
solid particles without the need for empirical drag force models.  Both
strategies are applicable in both meshless \cite{cleary2015prediction,
  trujillo2020smooth} and mesh-based \cite{van2008numerical, ma2022review}
CFD-DEM techniques.  In addition to the study mono-sized particle behaviour in
a fluid \cite{trujillo2020smooth}, research such as
\cite{brosh2014accelerating} and \cite{peng2021fully} explore wide particle
size distributions with CFD-DEM and SPH-DEM respectively.  Regarding
non-Newtonian fluid modelling in mesh-based schemes, works like
\cite{li2018dam} are notable and in meshless schemes, \cite{peng2021fully} is
notable. \citet{zhu2008discrete} and \citet{ma2022review} delve into
investigations on spherical and non-spherical particles respectively within
CFD-DEM frameworks.  Meshless CFD-DEM, though fruitful, encounters challenges
in handling problems with free surfaces and becomes computationally demanding
with a large number of solid particles, especially in fully resolved
coupling. In contrast, SPH offers advantages in such scenarios due to its
inherent capability in handling free surfaces and adeptness in modeling large
mesh deformation problems, facilitating the incorporation of novel physics.

The SPH method is a meshless numerical method originally proposed by
\citet{gingold1977smoothed} and \citet{lucy1977numerical} to model
astrophysics problems. It has been extensively applied to simulate problems
involving fluids, structural dynamics, fluid-structure interaction, granular
physics, non-Newtonian fluid flows \cite{peng2021fully} among other areas
\cite{monaghan2012smoothed}.  Various kind of particulate flows, involving,
simple spherical particles mixing using unresolved coupling
\cite{cleary2015prediction, markauskas2019coupled} and with debris flow
\cite{canelas2016sph} with fully resolved coupling approach were analysed
using SPH-DEM.  Handling particles of arbitrary shape in fluid flow is
addressed in various works, including those by \citet{peng2021fully,
  canelas2016sph, amicarelli2015smoothed}.  Several variants of SPH, such as
WCSPH \cite{peng2021fully, cleary2015prediction}, and ISPH
\cite{asai2021fluid}, are utilized for modeling fluid flow with particulate
flows. The coupling between fluid and rigid bodies in the fully resolved
category employs techniques such as the fixed ghost particle technique
\cite{canelas2016sph, asai2021fluid}, single layers of dummy SPH particles
\cite{peng2021fully}, and simple repulsive forces
\cite{monaghan2009sph}. Coupling between SPH fluid particles and solid
structures discretized as vertex edges is also proposed by
\citet{park2023new}. Particle settling with variable sizes is studied by
\citet{zou2022study}.  Elastic behavior of spherical particles and dispersion
studies are conducted by \citet{ng2021numerical}.  Interaction between rigid
bodies of arbitrary sizes is addressed using various techniques, such as those
discussed in papers by \citet{asai2021fluid}, \citet{peng2021fully},
\citet{canelas2016sph}.

In the current work we consider the particle dispersion within a large-scale
under stirrer mixing phenomenon. We employ Smoothed Particle Hydrodynamics
(SPH) to model fluid dynamics and the Discrete Element Method (DEM) to handle
the dynamics and contact interactions among spherical particles. The
interaction between solids and fluids is modelled using a fixed ghost particle
approach, ensuring a fully resolved simulation. We validate the fluid solver
independently using Poiseuille flow and the DEM solver separately using
fundamental benchmarks such as particle-particle and particle-wall
impacts. Subsequently, the fully resolved coupled SPH-DEM solver is validated
using three benchmarks: a single particle entering, two particles settling,
and a forced wedge entry into a steady hydrostatic tank. Upon confirming the
fidelity of the solver, we proceed with analyzing the mixing behavior of
various spherical particles influenced by a stirrer in a tank. We investigate
scenarios involving particles of different densities, stirrer velocities, and
variable sizes as part of our case study. The development work is conducted
using PySPH\cite{ramachandran2021pysph}, an open-source code available at
\url{https://github.com/pypr/pysph}.  We modify the current codebase to
integrate implementations of the Discrete Element Method (DEM) and the
Smoothed Particle Hydrodynamics (SPH) coupled with DEM model for our ongoing
research. To ensure reproducibility, we utilize the automan package
\cite{ramachandran2018automan} to automate all results generated in the
current manuscript, aiming for a reproducible research approach.

The structure of this paper is as follows: In
\cref{sec:fluid-modeling}, we explain the numerical method used to model fluid
dynamics. \Cref{sec:rbd} outlines the equations governing the dynamics
of rigid bodies and the contact force model employed to resolve collisions
among them. The coupling between rigid spherical particles and the fluid is
detailed in \cref{sec:rfc}. Our results are presented in
\cref{sec:results}, where various problems from the literature are simulated to
validate the current solver, and particle dispersion is examined.

\FloatBarrier%
\section{Fluid modeling}
\label{sec:fluid-modeling}
In the current work, we follow a weakly-compressible SPH approach to model the
fluid. The continuum in SPH is modeled using particles, which has physical
properties such as mass, velocity, and these particles interact based on the
governing equations using a Guassian-like kernel
\cite{monaghan-review:2005,morris1997modeling}.

\FloatBarrier%
\subsection{Fluid governing equations}
\label{sec:fluid--governing-equations}
The fluid motion is governed by the conservation of mass and the conservation
of linear momentum, which can be expressed by the following equations:
\begin{equation}
  \label{eq:ce}
  \frac{d \rho}{d t} = - \rho \; \frac{\partial u_i}{\partial x_i},
\end{equation}
\begin{equation}
  \label{eq:me}
  \frac{d u_i}{d t} = \frac{1}{\rho} \; \frac{\partial \sigma_{ij}}{\partial x_j}
  + g_i,
\end{equation}
where $\rho$ is the density, $u_i$ is the $i$\textsuperscript{th} component of
the velocity field, $x_j$ is the $j$\textsuperscript{th} component of the
position vector, $g_i$ is the component of body force per unit mass and
$\sigma_{ij}$ is stress tensor.

The stress tensor is split into pressure and viscous parts,
\begin{equation}
  \label{eq:fluid-stress-decomposition}
  \sigma_{ij} = - p \delta_{ij} + 2 \eta \frac{\partial u_i}{\partial x_j}
\end{equation}
where $\eta$ is the kinematic viscosity of the fluid, $p$ is the pressure,
$\delta_{ij}$ is the Kronecker delta function.

\FloatBarrier%
\subsection{Discretized fluid governing equations}
\label{sec:sph--governing-equations}
The governing equations involve function, derivative and divergence
operators. In SPH, these operators are approximated based on the positions,
mass and the kernel values of the discretized particles. Assume that the
domain is discretized into N particles with mass $m_i$ and volume
$V_i$, We have
\begin{equation}
  \label{eq:mass_repr}
  m_i = V_i \> \rho_i.
\end{equation}
Based on this discretization, the discrete function approximation is given as,
\begin{equation}
  \label{eq:discrete_form}
  A_i = \sum_{j \in \text{Neigh}(i)}\> \frac{m_j}{\rho_j} A_j\> W(\ten{x}_i - \ten{x}_j, h),
\end{equation}
where $A(\boldsymbol{x}_i) = A_i$ is the value of the field property of
particle i, similarly for $A_j$. Here, $\text{Neigh}(i)$ is the neighbours of
particle $i$.  A symmetric derivative approximation~\cite{Violeau16} is given
as,
\begin{equation}
  \nabla A(\ten{x}_i) = \rho_i \sum_{j} m_j \left(\frac{A_i}{\rho_i^2} + \frac{A_j}{\rho_j^2}\right) \nabla W_{ij}.
\end{equation}
A symmetric gradient operator ensures equal and opposite forces acting on the
particles.

An approximation of the divergence operator is used to discretize the
\cref{eq:ce}.  It is given as,
\begin{equation}
  \label{intro:eq:sph-continu-div-final}
  \nabla \cdot \ten{v}(\ten{x}_i) = \frac{1}{\rho_i} \sum_{j} m_j \left(\ten{v}_j - \ten{v}_i\right) \cdot \nabla W_{ij}.
\end{equation}

% \subsection{Discretized fluid governing equations}
The SPH discretization of the continuity
equation~\cref{eq:sph-discretization-continuity} is given as,
\begin{equation}
  \label{eq:sph-discretization-continuity}
  \frac{{d}\rho_a}{dt} = \sum_{b} \; \frac{m_b}{\rho_{b}} \;
  \rho_{a} \; {\ten{u}}_{ab} \; \cdot \nabla_{a} W_{ab},
\end{equation}

%
Similarly, the discretized momentum equation is written as,
\begin{multline}
  \label{eq:sph-momentum-fluid}
  \frac{{d}\ten{u}_{a}}{dt} = - \sum_{b} m_b
  \bigg(\frac{p_a}{\rho_a^2} + \frac{p_b}{\rho_b^2}\bigg)
  \nabla_{a} W_{ab}
 \;+\;
  \sum_{b} m_b \frac{4 \eta \nabla W_{ab}\cdot
    \ten{r}_{ab}}{(\rho_a + \rho_b) (r_{ab}^2 + 0.01 h_{ab}^2)} \ten{u}_{ab}  \;+\;
  \ten{g}_{a},
\end{multline}
where $\ten{I}$ is the identity matrix, $\eta$ is the kinematic viscosity of the
fluid. The viscous term in \Cref{eq:me} is discretized according to the
formulation introduced in \cite{morris1997modeling}.

We add to the momentum equation an additional artificial viscosity term
$\Pi_{ab}$~\cite{monaghan-review:2005} to maintain the stability of the
numerical scheme, given as,
\begin{align}
  \label{eq:mom-av}
  \Pi_{ab} =
  \begin{cases}
\frac{-\alpha h_{ab} \bar{c}_{ab} \phi_{ab}}{\bar{\rho}_{ab}}
  & \ten{u}_{ab}\cdot \ten{r}_{ab} < 0, \\
  0 & \ten{u}_{ab}\cdot \ten{r}_{ab} \ge 0,
\end{cases}
\end{align}
where,
%
\begin{equation}
  \label{eq:av-phiij}
  \phi_{ab} = \frac{\ten{u}_{ab} \cdot \ten{r}_{ab}}{r^2_{ab} + 0.01 h^2_{ab}},
\end{equation}
%
where $\ten{r}_{ab} = \ten{r}_a - \ten{r}_b$,
$\ten{u}_{ab} = \ten{u}_a - \ten{u}_b$, $h_{ab} = (h_a + h_b)/2$,
$\bar{\rho}_{ab} = (\rho_a + \rho_b)/2$, $\bar{c}_{ab} = (c_a + c_b) / 2$, and
$\alpha$ is the artificial viscosity parameter.  The pressure $p_a$ is evaluated
using an equation of state:
\begin{equation}
\label{eqn:sph-eos}
  p_a = K \bigg(\frac{\rho_a}{\rho_{0}} - 1 \bigg).
\end{equation}
Where, $K=\rho_0 \, c_0^2$ is bulk modulus of the body, with
$c_0=10 \times V_{\text{max}}$ is speed of sound, while $\rho_0$ as the
initial density of the particles.

\FloatBarrier%
\subsection{Boundary Conditions}
\label{sec:boundary_conditions}

The ghost particle approach proposed by \citet{Adami2012} is used to model the
boundaries. We use three layers of ghost particles to model the solid wall.
The properties of the solid wall are interpolated from the fluid particles.

When the viscous force is computed, the no slip boundary condition is used,
where the velocity on the boundary set as,
\begin{equation}
  \label{eq:no-slip-bc-u}
  \ten{u}_{\text{Ga}} = 2 \ten{u}_{\text{p}} - \ten{\hat{u}}_{\text{a}}.
\end{equation}
The projected velocity $\ten{\hat{u}}_{\text{a}}$ on the ghost particles is
computed using,
\begin{equation}
  \label{eq:v-ghost}
  \ten{\hat{u}}_a = \frac{\sum_b\ten{u}_b W_{ab}}{\sum_b W_{ab}},
\end{equation}
where $\ten{u}_b$, is the velocity of the fluid particle $b$ and $W_{ab}$ is the kernel
value between the fluid particle and the ghost particle.

The pressure of the boundary particle is extrapolated from its surrounding
fluid particles by the following equation,
\begin{equation}
  \label{eq:pressure-bc}
  p_w = \frac{\Sigma_f p_f W_{wf} + (\ten{g} - \ten{a}_{\ten{w}}) \cdot \Sigma_f
    \rho_f \ten{r}_{wf} W_{wf}}{\Sigma_f W_{wf}},
\end{equation}
where $\ten{a}_w$ is the acceleration of the wall. The subscript $f$ denotes
the fluid particles and $w$ denotes the wall particles.

\FloatBarrier%
\section{Rigid body dynamics}
\label{sec:rbd}
% The rigid body is discretized into particles with equal spacing each particle
% with mass $m_i$ and density $\rho_i$. Rigid body has a total 6 degrees of
% freedom (DOF), divided into $3$ translational and $3$ rotational.

@@

\begin{figure}[!htpb]
  \centering
  \includegraphics[width=0.7\textwidth]{images/rigid_body/rigid_body}
  \caption{Body frame and local frame description of rigid body}
  \label{fig:gloabl_body_frame_rb}
\end{figure}
We use two coordinate frames to capture the dynamics of the rigid body, a
global frame and a body frame as shown in
\cref{fig:gloabl_body_frame_rb}. The body fixed frame, which moves with
rigid body is always located at the center of mass ($\ten{x}_{cm}$). The
state of the rigid body at a given time ($t$) can be described using position
($\ten{x}_{cm}$) and velocity ($\ten{v}_{cm}$) of the center of mass, a
rotation matrix($\ten{R}$) to represent the orientation of the rigid body with
respect to the global frame, and angular velocity($\teng{\omega}$). The center
of mass is computed as
\begin{equation}
  \label{eq:rfc:center_of_mass}
  \ten{x}_{cm} = \frac{\sum_i m_i \; \ten{x}_{i} }{\sum_i m_i }.
\end{equation}
The position of the discretized particle ($i$) in
\cref{fig:gloabl_body_frame_rb} belonging to the rigid body at time $t$ can be
computed as,
\begin{equation}
  \label{eq:rfc:rb_particle_pos_update}
  \ten{x}_i = \ten{x}_{cm} + \ten{r}_{i},
\end{equation}
with
\begin{equation}
  \label{eq:rfc:rb_particle_pos_update}
  \ten{r}_i = \ten{R} \overline{\ten{r}}_{i}.
\end{equation}
Here $\overline{\ten{r}}_{i}$ is the position of the particle $i$ about the body
frame axis and remains constant through out the simulation. The rotation matrix
$\ten{R}$ is used to bring the body frame position vector to the global frame
$\ten{O}$. Similarly the velocity vector is computed as,
\begin{equation}
  \label{eq:rfc:rb_particle_vel_update}
  \ten{v}_i = \ten{v}_{cm} + \teng{\omega} \times \ten{r}_{i}.
\end{equation}

We evolve the state of the rigid body through the integration of the
\cref{eq:rfc:balance_linear_mom,eq:rfc:balance_angular_mom}. The linear velocity of the
center of mass ($\ten{v}_{cm}$) and angular momentum ($\ten{L}$) at the next
timestep are computed as,
\begin{equation}
  \label{eq:rfc:lin_vel_cm_update}
  \ten{v}_{cm}^{n+1} = \ten{v}_{cm}^{n} + \frac{\ten{F}_{cm}}{M} \; \Delta t,
\end{equation}
\begin{equation}
  \label{eq:rfc:ang_mom_update}
  \ten{L}^{n+1} = \ten{L}^{n} + \teng{\tau}_{cm} \; \Delta t.
\end{equation}
Here, $\ten{F}_{cm} = \sum_i \ten{F}_i$.

The position of the center of mass and the rotation matrix ($\ten{R}$) are updated
by,
\begin{equation}
  \label{eq:rfc:lin_pos_cm_update}
  \ten{x}_{cm}^{n+1} = \ten{x}_{cm}^{n} + \ten{v}_{cm}^{n} \; \Delta t,\\
  \ten{R}^{n+1} = \ten{R}^{n} + \tilde{\teng{\omega}}^{n} \, \ten{R}^{n} \; \Delta t,
\end{equation}
where $\tilde{\teng{\omega}}^{n}$ is matrix formulation of angular velocity
$\omega$. The angular velocity at the new time step is computed with
\begin{equation}
  \label{eq:rfc:ang_velocity_update}
  \teng{\omega}^{n+1} = (\textit{\teng{I}}^{-1})^{n+1} \; \ten{L}^{n+1}.
\end{equation}
Here, moment of inertia at the new time step is computed as,
\begin{equation}
  \label{eq:rfc:moi_update}
  (\textit{\teng{I}}^{-1})^{n+1} = \ten{R}^{n+1} \textit{\teng{\overline{I}}}^{-1} (\ten{R}^{n+1})^T.
\end{equation}
where moment of inertia ($\textit{\teng{\overline{I}}}^{-1}$) in body frame is
used to compute in global frame at every time instant for faster computations.
The moment of inertia ($\textit{\teng{\overline{I}}}$) is computed as,
\begin{equation*}
\textit{\teng{\overline{I}}} =
\begin{bmatrix}
\sum_i m_i (y_i^2 + z_i^2) & -\sum_i m_i x_iy_i & -\sum_i m_i x_iz_i\\
-\sum_i m_i x_iy_i & \sum_i m_i (x_i^2 + z_i^2) &  -\sum_i m_i y_iz_i\\
-\sum_i m_i  x_iz_i & -\sum_i m_i y_iz_i & \sum_i m_i (x_i^2 + y_i^2)
\end{bmatrix}.
\end{equation*}

The position and velocity of the particles of the rigid body are updated by
\begin{eqnarray}
  \label{eq:rfc:rb_particle_pos_update}
  \ten{r}_i = \ten{R} \cdot \overline{\ten{r}}_{i},\\
  \ten{x}_i = \ten{x}_{cm} + \ten{r}_{i},\\
  \ten{v}_i = \ten{v}_{cm} + \teng{\omega} \times \ten{r}_{i}.
\end{eqnarray}

>>>>

\documentclass[a4paper,11pt]{book}

% TEMPLATE
% ----------------------------------------------------------------------
\usepackage[utf8]{inputenc}
\usepackage{geometry}
\geometry{
	a4paper,
	left=20mm,
	top=30mm,
	bottom=30mm,
}
\usepackage[hidelinks]{hyperref}
\usepackage{titlesec}
\usepackage{etoolbox}
\usepackage{chngcntr}
\usepackage{graphicx}
\usepackage{lipsum}
\usepackage{xcolor}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{fancyhdr}
\usepackage{booktabs}

\counterwithout{figure}{chapter}
\renewcommand\thesection{\arabic{section}}
\patchcmd{\thebibliography}{\chapter*}{\section*}{}{}
\renewcommand{\bibname}{References}
\bibliographystyle{plainnat}
\pagestyle{empty}
\titleformat{\chapter}[frame]{\normalfont}{}{10pt}{\LARGE\bfseries\filcenter}
\pagestyle{fancy}
\fancyhf{}
\pagestyle{fancy}
% \fancyhead[LE,RO]{}
% \fancyhead[RE,LO]{Modelling for Engineering \& Human Behaviour 2024}
% \fancyfoot[C]{\thepage}

\renewcommand{\theequation}{\arabic{equation}}
\renewcommand{\thefigure}{\arabic{figure}}
\renewcommand{\thetable}{\arabic{table}}
% ----------------------------------------------------------------------

% ==============================
% Imports from my paper
% ==============================
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{multirow}
\usepackage{relsize}
\usepackage[utf8]{inputenc}
\usepackage[capitalise]{cleveref}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\usepackage[section]{placeins}
\usepackage{booktabs}
\usepackage{url}

@@

\usepackage{lineno}
\usepackage{natbib}
% ==============================
% Imports from my paper ends
% ==============================

@@

\chapter{{DEM}-{SPH} study of particles dispersion in fluid}

\thispagestyle{empty}
\vspace{-0.5cm}
\begin{center}
\begin{large}
  \underline{Dinesh Adepu}$^\flat$\footnote{d.adepu@surrey.ac.uk}
  and Chuan-Yu Wu$^\flat$\\
  % and C. Author$^\flat$\\
  \vspace{0.25cm}
  \normalsize ($\flat$)\; School of Chemistry and Chemical Engineering, University of Surrey\\
  \normalsize  Guildford, Surrey, United Kingdom.\\
  % \vspace{0.1cm}
  % \normalsize ($\natural$)\;  Second Author Affiliation,\\
  % \normalsize Second Author Address.\\
\end{large}
\end{center}
\vspace{-0.5cm}

\section{Introduction}

Particle dispersion is crucial in various industries from food processing to
ice-sea modeling where powder and fluids interact.  Studying particle dispersion
through experiments can be difficult due to the complex interactions involved.
Numerical methods, however, offer a powerful alternative, allowing researchers
to explore these systems using mesh-based or meshless techniques.  Computational
Fluid Dynamics and Discrete Element Method (CFD-DEM) is a common approach for
simulating particle flow.  Mesh-based methods like lattice Boltzmann and finite
volume methods handle the fluid, while meshless methods like Smoothed Particle
Hydrodynamics (SPH) handle it without a mesh. Mesh based methods struggle with
free surfaces and large particle numbers, while meshless methods excels in these
areas. Among many meshless techniques, SPH has been used for various particulate
flows, including mixing of simple spheres and debris flow with complex shapes.

This work simulates particle dispersion in a stirred tank using SPH for fluids
and DEM for particles.  The interaction between particles and fluids is fully
resolved.  We validate both the SPH and DEM solvers individually, followed by
validation of the combined SPH-DEM solver.  Then we analyze how different
particle properties and stirrer velocities affect mixing behavior. We use an
open-source code called PySPH and modify it. We use an automation package to
ensure our results are reproducible.

\section{Modeling}
\subsection{Fluid modeling}

% \subsection{Discretized fluid governing equations}
In the current work, we follow a weakly-compressible SPH approach to model the
fluid. The continuum in SPH is modeled using particles, which has physical
properties such as mass, velocity, and these particles interact based on the
governing equations using a Guassian-like kernel. The SPH discretized governing
equations for the continuity equation and the momentum equation of a fluid are
given as:
\begin{equation}
  \label{eq:sph-discretization-continuity}
  \frac{{d}\rho_a}{dt} = \sum_{b} \; \frac{m_b}{\rho_{b}} \;
  \rho_{a} \; {\ten{u}}_{ab} \; \cdot \nabla_{a} W_{ab},
\end{equation}
\begin{multline}
  \label{eq:sph-momentum-fluid}
  \frac{{d}\ten{u}_{a}}{dt} = - \sum_{b} m_b
  \bigg(\frac{p_a}{\rho_a^2} + \frac{p_b}{\rho_b^2} + \Pi_{ab}\bigg)
  \nabla_{a} W_{ab}
 \;+\;
  \sum_{b} m_b \frac{4 \eta \nabla W_{ab}\cdot
    \ten{r}_{ab}}{(\rho_a + \rho_b) (r_{ab}^2 + 0.01 h_{ab}^2)} \ten{u}_{ab}  \;+\;
  \ten{g}_{a},
\end{multline}
where $\rho$ is the density, $\ten{u}$ is the velocity, $g$ is the component of
body force per unit mass, where $\eta$ is the kinematic viscosity of the fluid,
$p$ is the pressure.  Artificial viscosity term $\Pi_{ab}$, given as,
\begin{align}
  \label{eq:mom-av}
  \Pi_{ab} =
  \begin{cases}
\frac{-\alpha h_{ab} \bar{c}_{ab} \phi_{ab}}{\bar{\rho}_{ab}}
  & \ten{u}_{ab}\cdot \ten{r}_{ab} < 0, \\
  0 & \ten{u}_{ab}\cdot \ten{r}_{ab} \ge 0,
\end{cases}
\end{align}
where,
%
\begin{equation}
  \label{eq:av-phiij}
  \phi_{ab} = \frac{\ten{u}_{ab} \cdot \ten{r}_{ab}}{r^2_{ab} + 0.01 h^2_{ab}},
\end{equation}
%
where $\ten{r}_{ab} = \ten{r}_a - \ten{r}_b$,
$\ten{u}_{ab} = \ten{u}_a - \ten{u}_b$, $h_{ab} = (h_a + h_b)/2$,
$\bar{\rho}_{ab} = (\rho_a + \rho_b)/2$, $\bar{c}_{ab} = (c_a + c_b) / 2$, and
$\alpha$ is the artificial viscosity parameter.  The pressure $p_a$ is evaluated
using an equation of state:
\begin{equation}
\label{eqn:sph-eos}
  p_a = K \bigg(\frac{\rho_a}{\rho_{0}} - 1 \bigg).
\end{equation}
Where, $K=\rho_0 \, c_0^2$ is bulk modulus of the body, with
$c_0=10 \times V_{\text{max}}$ is speed of sound, while $\rho_0$ as the
initial density of the particles.

\FloatBarrier%
\subsection{Rigid body dynamics}
\label{sec:rbd}
% The rigid body is discretized into particles with equal spacing each particle
% with mass $m_i$ and density $\rho_i$. Rigid body has a total 6 degrees of
% freedom (DOF), divided into $3$ translational and $3$ rotational.

@@

The force acting on particle $i$ is composed of interaction with the other rigid
bodies, and the fluid, given as
\begin{eqnarray}
  \label{eq:rfc:rb_particle_pos_update}
  \ten{F}_i = \ten{F}_{\text{Fl}}^i + \ten{F}_{\text{cont}}^i
\end{eqnarray}
We follow DEM to compute force $\ten{F}_{\text{cont}}^a$ acting on particle $i$
due to the interaction with the rigid bodies. The force $\ten{F}_{\text{rfc}}^i$
acting due to the interaction with the fluid particles is computed by resolving
the rigid body into dummy SPH particles.  These SPH particles are evenly
distributed and remain stationary, moving in tandem with the velocity of the
rigid particle at any given location.  They serve as SPH boundary particles and
contribute to the computation of fluid particle density and acceleration.  The
force acting on the sampled dummy SPH particle due to the interaction with the
fluid is given by,
\begin{equation}
  \label{eq:rfc-force}
  \ten{F}_{\text{rfc}}^a = -m_a \sum_{f} m_f \bigg(\frac{p_f}{\rho_{f}^2} +
  \frac{p_a}{\rho_{a}^2}\bigg) \nabla_{a} W(x_{af}) +
  m_a \sum_{f} m_f \frac{4 \eta \nabla_a W_{af}\cdot
    \ten{r}_{af}}{(\rho_a + \rho_f) (r_{af}^2 + 0.01 h_{af}^2)} \ten{u}_{af}
\end{equation}
where, $m_a$ signifies the hydrodynamic mass of the sampled dummy SPH particle,
and $\rho_a$ represents its hydrodynamic density. While $m_f$, $p_f$ and
$\rho_f$ are mass, pressure and density of the fluid particle.

\section{Results}

We validate the Discrete Element Method solver by simulating a particle-wall
impact scenario.  We validate the coupled SPH-DEM solver with simulations
involved like a circular particle entering a steady tank.  Lastly, we
investigate the mixing behavior of spherical particles under a stirrer,
examining cases with varying particle diameters and stirrer speeds.

\FloatBarrier%
\subsection{DEM validation: Particle wall impact}
\label{sec:DEM_validation_2_particle_wall_impact}
\begin{figure}[!htpb]
  \centering
  \begin{subfigure}{0.48\textwidth}
    \centering
    \includegraphics[width=0.6\textwidth]{images/results_dem_2_validation_particle_wall_impact/schematic}
    \subcaption{
    }\label{fig:result:dem-1a}
  \end{subfigure}
  \begin{subfigure}{0.48\textwidth}
    \centering
    \includegraphics[width=0.6\textwidth]{figures/benchmark_4_sw_colliding_oblique/angle_vs_ang_vel.pdf}
    \subcaption{
    }\label{fig:result:dem-1b}
  \end{subfigure}
  \caption{(a) Schematic of particle-wall impact.  (b)
 Variation of Angular velocity with impact angle.}
\label{fig:result-dem-1}
\end{figure}
We consider the collision of a spherical particle with wall with different
incident angles and a constant velocity in magnitude. This test case is useful
in testing the tangential interaction modeling of our DEM solver.  The schematic
of the body and wall is shown in \cref{fig:result:dem-1a}.  The radius of the
impacting particle is $2.5 \times 10^{-3}$ m and is made of aluminium oxide. The
particle has a Young's modulus of $3.8 \times 10^{11}$ N\,m\textsuperscript{-2}, a
Poisson's ratio of $0.23$, and a density of $4,000$
kg\,m\textsuperscript{-3}. The wall's Young's modulus is $70\times 10^{9}$
N\,m\textsuperscript{-2}, with a Poisson's ratio of $0.25$.  The impact velocity
has a magnitude of $3.9$ m\,s\textsuperscript{-1}. The coefficient of friction
between the wall and the particle is $0.092$.  In
\cref{fig:result:dem-1b}, we show variation of the
rebound angular velocity with the incident impact angle, compared to
experimental data.

\FloatBarrier%
\subsection{Single particle entering into a tank}
\label{sec:rfc_validation_1_single_particle_entry}
\begin{figure}[!htpb]
  \centering
  \begin{subfigure}{0.48\textwidth}
    \centering
    \includegraphics[width=0.6\textwidth]{
      images/rfc_01_skillen_2013_particle_entry_in_hs_tank/Skillen_2013_particle_entry_in_hs_tank}
    \subcaption{}\label{fig:result:rfc-1a}
  \end{subfigure}
  \begin{subfigure}{0.48\textwidth}
    \centering
    \includegraphics[width=0.6\textwidth]{figures/skillen_2013_water_entry_half_buoyant/penetration_vs_t}
    \subcaption{}\label{fig:result:rfc-1b}
  \end{subfigure}
  \caption{(a) Schematic of cylinder disc entering a 2D tank.  (b) Evolution of
    the penetration depth of the cylinder disc with time. }
\label{fig:result-rfc-1}
\end{figure}
In this example we examine the motion of a circular disc descending into a
steady hydrostatic tank. \Cref{fig:result:rfc-1a} illustrates the initial setup,
with the cylinder positioned $0.5$ meters from the free surface. The cylinder's
radius is $0.11$ meters, and density of $500$ kg\,m\textsuperscript{-3}.
\Cref{fig:result:rfc-1b}, depicts the evolution of the depth of penetration of
the cylinder compared to experimental results and other numerical techniques.

% \FloatBarrier%
\subsection{Mixing of spherical particles in a fluid tank}
\label{sec:mixing-spherical-particles-in-fluid-tank-homogeneous}
In this section, we investigate particle dispersion within a tank under various
stirrer velocities. Each circular particle has a radius of $0.055$ m, with a
total of $54$ particles.  The density of the particles is taken as $1050$
kg\,m\textsuperscript{-3}.  After allowing the particles to settle for $1$
second, the stirrer oscillates throughout the fluid length. We examine three
stirrer speeds.  The schematic of the figure is given in
\Cref{fig:1-mixing-1-a}.  \Cref{fig:1-mixing-1-b,fig:1-mixing-1-c,fig:1-mixing-1-d},
illustrate the distribution of particles within a fluid tanker under the
influence of a stirrer at different instants. From \Cref{fig:1-mixing-1},
particles tend to aggregate and settle gradually into a clump at the center.
